---
title: "Time Series Graphics"
subtitle: "Chapter 2"
format: md
editor: visual
project:
  execute-dir: project
---

## 2.2 Time Plots

### Setup

```{r}
#| label: setup
#| message: true
#| warning: false
#| paged-print: false
library(tidyverse)
library(tsibble)
library(fpp3)
library(ggtime)
```

### Simple graph

```{r}
head(ansett) #passenger numbers on Ansett airline flights

melsyd_economy <- ansett |> 
  filter(Airports == "MEL-SYD", Class=="Economy") |> 
  mutate(Passengers = Passengers/1000) 

autoplot(melsyd_economy) + 
  labs(
    title="Ansett arilines economy class",
    subtitle="Melbourne-Sydney",
    y="Passengers (k)"
  )


```

```{r}
head(PBS)

a10 <- PBS |>
  filter(ATC2 == "A10") |>
  select(Month, Concession, Type, Cost) |>
  summarise(TotalC = sum(Cost)) |>
  mutate(Cost = TotalC / 1e6) -> a10

autoplot(a10, Cost) +
  labs(y="$ (millions)",
       title = "Australian antidiabetic drug sales")
```

## 2.4 Seasonal Plot

```{r}
a10 |> 
  gg_season(Cost, labels="both") +
  labs(y="$M", title="Seasonal plot: Antidiabetic drug sales")
```

```{r}
head(vic_elec)

vic_elec |> 
  gg_season(Demand, period="week") +
  theme(legend.position = "none") +
  labs(y="MWh", title="Electicity demand: Victoria")
```

### 2.5 Seasonal subqueries plot

```{r}
a10 |> 
  gg_subseries(Cost) +
  labs(y="$ M", title="Australian antidiabetic drug sales")
```

### Example: Australian Holiday Tourism

```{r}
head(tourism)

holidays <- tourism |> 
  filter(Purpose=="Holiday") |> 
  group_by(State) |> 
  summarise(Trips=sum(Trips))

head(holidays)

autoplot(holidays, Trips) +
  labs(
    y="Overnight trips (k)",
    title="Australian Domestic Holidays"
  )
```

To see the timing of the seasonal peaks in each state, we can use a season plot. Figure [2.10](https://otexts.com/fpp3/subseries.html#fig:holidaysseason) makes it clear that the southern states of Australia (Tasmania, Victoria and South Australia) have strongest tourism in Q1 (their summer), while the northern states (Queensland and the Northern Territory) have the strongest tourism in Q3 (their dry season).

```{r}
gg_season(holidays, Trips) +
  labs(y = "Overnight trips ('000)",
       title = "Australian domestic holidays")
```

```{r}
holidays |> 
  gg_subseries(Trips)
```

### 2.6 Scatter Plot

```{r}
head(vic_elec)

vic_elec |> 
  filter(year(Time)==2014) |> 
  autoplot(Demand) +
  labs(y="GW", title="Half-hour eletricity demand: Victoria")

vic_elec |> 
  filter(year(Time)==2014) |> 
  autoplot(Temperature) +
  labs(y="°C", title="Half-hour eletricity demand: Victoria")

vic_elec |> 
  filter(year(Time)==2014) |> 
  ggplot(aes(x=Temperature, y=Demand)) +
  geom_point(alpha=.5, colour="darkblue") +
  theme_light() +
  labs(x="°C", y="GW")

```

#### Scatter Plot Matrices

```{r}
head(tourism)

visitors <- tourism |> 
  group_by(State) |> 
  summarise(Trips=sum(Trips))

visitors |> 
  ggplot(aes(x=Quarter, y=Trips)) +
  geom_line() +
  facet_grid(vars(State), scales="free_y") +
  labs(title = "Australian domestic tourism",
       y= "Overnight trips (k)")
```

```{r}
#| message: false
visitors |> 
  pivot_wider(values_from=Trips, names_from = State) |> 
  GGally::ggpairs(columns=2:9)
```

## 2.7 Lag plots

```{r}
head(aus_production)

recent_production <- aus_production |> 
  filter(year(Quarter) >= 2000)

recent_production |> 
  gg_lag(Beer, geom = "point") +
  labs(x="lag(Beer, k)")
```

The relationship is strongly positive at lags 4 and 8, reflecting the strong seasonality in the data. The negative relationship seen for lags 2 and 6 ocurrs because peaks (in Q4) are plotted against trughs (in Q2)

## 2.8 Autocorrelation

Just as correlation measures the extent of a linear relationship between two variables, autocorrelation measures the linear relationship between *lagged values* of a time series.

```{r}
recent_production |> 
  ACF(Beer, lag_max=9)
```

The values in the ACF column are the "r" corresponding to the ninge scatterplots before.

```{r}
recent_production |> 
  ACF(Beer) |> 
  autoplot() +
  labs(title="Australian beer production")
```

Graphic visualization of r values, the dashed blue lines indicate whether the correlations are significantly different from zero.

### Trend and seasonality in ACF plots

When data have a trend, the autocorrelations for small lags tend to be large and positive because observations nearby in time are also nearby in value. So the ACF of a trended time series tends to have positive values that slowly decrease as the lags increase.

When data are seasonal, the autocorrelations will be larger for the seasonal lags (at multiples of the seasonal period) than for other lags.

When data are both trended and seasonal, you see a combination of these effects.

```{r}
a10 |> 
  ACF(Cost, lag_max=48) |> 
  autoplot() +
  labs(title="Australian antidiabetic drug sales")
```

## 2.9 White noise

Time series that show no autocorrelation are called white noise

```{r}
set.seed(30)
y <- tsibble(sample=1:50, wn=rnorm(50), index=sample)
y |> 
  autoplot(wn) +
  labs(title="white noise", y="")
```

```{r}
y |> 
  ACF(wn) |> 
  autoplot() +
  labs(title="White Noise")
```

### site quest: is dollar white noise?

```{r}
library(rbcb)
dollar <- get_series(1, start_date=dmy(25112020)) |> 
  set_names(c("date", "price")) |> 
  as_tsibble(index=date)

head(dollar)

dollar |> 
  autoplot() +
  labs(title="PTAX Series", y="R$", x="date")

```

```{r}
dollar |> 
  gg_lag(y=price, lags = 1:30, geom = "point")

dollar |> 
  mutate(index=row_number()) |> 
  as_tsibble(index=index) |> 
  ACF(lag_max = 3*365) |> 
  autoplot()
```

## 2.10 Exercises

### Number 3

```{r}

tute <- read_csv("../data/tute1.csv") |> 
  janitor::clean_names()
head(tute)

myts <- tute |> 
  mutate(quarter = yearquarter(quarter)) |> 
  as_tibble(index=quarter)
  
myts |> 
  pivot_longer(cols=-quarter) |> 
  ggplot(aes(x=quarter, y=value, color=name)) +
  geom_line() +
  facet_grid(name~., scales="free_y") +
  theme_minimal()
```
